---
title: "Network"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(igraph)
```

Our starting point is that while causal influence flows one way through the graph,
along the directions of arrows from parents to children, statistical information can
flow in either direction.

Want to measure the effect of a time varying treatment on an outcome.

Let's assume that our sample consists of HIV positive patients who are treated at two time points and that between the first and second treatment we have a measurement of CD4 cell count (high is good, indicates a good immune system). $U$ is the unmeasured immunosupression level. $Y$ is binary.


```{r}

# sort of interesting that this doesn't work very well.

set.seed(2)
# simplest case of TVC
n = 1000
A0 = rbinom(n, 1, p = .5)
U = rbinom(n, 1, p = .3) # U=1 people are super sick
L1 = rnorm(n, 100 + 10 * A0 - 20 * U) # CD4 cell count (low is bad)
A1 = rbinom(n, 1, p = plogis(.3*(100-L1)))
Y = rbinom(n, 1, prob = plogis(-1+U))

dat = data.frame(cbind(A0, L1, A1, Y))

m0 = (glm(Y ~ A0 + A1 + L1, family = binomial))
summary(m0)



m1 = (lm(L1 ~ A0))
#coefficients(m0)[4] * coef(m1)[1] + coefficients(m0)[1] # new intercept
#coefficients(m0)[4] * coef(m1)[2] + coefficients(m0)[2]

#plot(factor(A0), Y)

# weighted MSM

# first, come up with treatment weights
tm = glm(A1 ~ L1, data = dat, family = binomial)

wt = ifelse(dat$A1 == 1, sum(A1)/(n*predict(tm, type = "response")), (n-sum(A1))/( n * (1- predict(tm, type = "response")) ))



msm = glm(Y ~ A0 + A1, weights = wt, family = binomial, data = dat)
summary(msm)


```

$$
P(A_0,A_1 \vert L_1) = P(A_0) P(A_1 \vert L_1)
$$
P(A0, A1 | L1)


## R Markdown

A CGM describes how the structure of a system is modified by an intervention.

If it is possible to de- rive an expression for the desired post-interventional dis- tribution purely in terms of the joint distribution over the original system via the do-calculus then the causal query is identifiable.

Explicitly created a joint model over the system pre and post-intervention, which allows the direct application of standard statistical inference, rather than requiring additional notation and operations that map from one to the other.

To infer causal effects, compute a predictive distribution for the quantity of interest in the post-intervention graph using Bayes rule.
